An Aliasing-Tolerant Parallel Programming Model

### 2012-11-11 doc/types-pldi2013/langdef_full.tex
added appendix - did first pass of cleaning up label references

\newtheorem{thm}{Theorem}
\newtheorem{lem}{Lemma}

@@

% fun latex tricks to make typeenv and opsenv more friendly
\makeatletter
\define@key{typeenv}{G}{\def\typeenv@G{#1}}
\define@key{typeenv}{P}{\def\typeenv@P{#1}}
\define@key{typeenv}{O}{\def\typeenv@O{#1}}
\newcommand{\typeenvx}[1][]{
{
% default values
\def\typeenv@G{\Gamma}
\def\typeenv@P{\Phi}
\def\typeenv@O{\Omega}
\setkeys{typeenv}{#1}
\typeenv@G, \typeenv@P, \typeenv@O \vdash \,
}}
\newcommand{\typeenv}[3][]{\typeenvx[#1] {#2} : {#3}}
\define@key{opsenv}{M}{\def\opsenv@M{#1}}
\define@key{opsenv}{L}{\def\opsenv@L{#1}}
\define@key{opsenv}{H}{\def\opsenv@H{#1}}
\define@key{opsenv}{S}{\def\opsenv@S{#1}}
\define@key{opsenv}{C}{\def\opsenv@C{#1}}
\newcommand{\opsenvx}[1][]{
{
% default values
\def\opsenv@M{M}
\def\opsenv@L{L}
\def\opsenv@H{H}
\def\opsenv@S{S}
\def\opsenv@C{C}
\setkeys{opsenv}{#1}
\opsenv@M, \opsenv@L, \opsenv@H, \opsenv@S, \opsenv@C \vdash \,
}}
\newcommand{\opsenv}[4][]{\opsenvx[#1] {#2} \mapsto {#3}, {#4}}
\makeatother

\begin{figure*}
\centering
{\small
\begin{tabular}{cclr|cclr}

$T$ & ::= &  & types & $bv$ & ::= & false $\;\;\;\mid\;\;\;$ true & \\
  &$\mid$& bool $\;\;\;\mid\;\;\;$ int & base types & & & & \\
  &$\mid$& $\langle T_1, T_2 \rangle$ & tuple & $iv$ & ::= & 0 $\;\;\;\mid\;\;\;$ 1 $\ldots$ & \\
  &$\mid$& $T@(\oton{r})$ & pointer & & & & \\
  &$\mid$& $\text{coloring}(r)$ & region coloring & $e$ & ::= & & expressions \\
  &$\mid$& $\exists \oton{r}. T\text{ where }\Omega$ & region relationship &   &$\mid$& $bv$ $\;\;\;\mid\;\;\;$ $iv$ & constants \\
  &$\mid$& $\forall \oton{r}. (\oton{T}), \Phi, Q \rightarrow T_r$ & functions &   &$\mid$& $\langle e_1, e_2 \rangle$ $\;\;\;\mid\;\;\;$ $e$.1 $\;\;\;\mid\;\;\;$ $e$.2 & tuple \\
& & & &   &$\mid$& $id$ &  \\
$\Omega$ & ::= & $\{ \oton{\omega} \}$ & region constraints &   &$\mid$& $\text{new}\ T@r$ $\;\;\;\mid\;\;\;$ $\text{null }T@r$ $\;\;\;\mid\;\;\;$ $\text{isnull}(e)$ & \\
$\omega$ & ::= & $r_1 \leq r_2$ & subregion   &   
  &$\mid$& $\text{upregion}(e, r_1,\ldots,r_n)$ & \\
  &$\mid$& $r_1 * r_2$ & disjointness &   
  &$\mid$ & $\text{downregion}(e, r_1,\ldots,r_n)$ & \\  
& & &  & 
  &$\mid$& $\text{read}(e_1)$ $\;\mid\;$ $\text{write}(e_1, e_2)$ & memory access \\
$\Phi$ & ::= & $\{ \oton{\phi} \}$ & privileges & 
  & $\mid$ & $\text{reduce}(id, e_1, e_2)$ & \\
$\phi$ & ::= & reads$(r)$ $\;\mid\;$ writes$(r)$ $\;\mid\;$ $\text{reduces}_{id}(r)$ &  &  
  &$\mid$& $\text{newcolor}\ r$ $\;\mid\;$ $\text{color}(e_1, e_2, e_3)$ & coloring \\
& & & &  
  &   $\mid$& $e_1 + e_2$ & integer ops \\
$Q$ & ::= & $\{ \oton{q} \}$ & coherence modes  &   
       & $\mid$& $e_1 < e_2$ & comparisons \\
$q$ & ::= & atomic$(r)$ $\;\;\;\mid\;\;\;$ simult$(r)$ & &   
  &$\mid$& $\text{let}\ id : T = e_1 \text{in}\ e_2$ &  \\
& & & &   
       & $\mid$& $\text{if}\ e_1\ \text{then}\ e_2\ \text{else}\ e_3$ &  \\
$v$ & ::= & & values &   
       & $\mid$& $id[r_1, \ldots, r_n](e_1,\ldots,e_n)$ & function calls \\
  &$\mid$& $bv$ $\;\;\;\mid\;\;\;$ $iv$ & base values &   
       & $\mid$ & $\text{partition}\ r_p\text{ using }e_1\text{ as }\oton{r}\text{ in }\ e_2$ &  \\
  &$\mid$& $\langle v_1, v_2 \rangle$ & tuple & 
       & $\mid$& $\text{pack}\ e_1\ \text{as}\ T[r_1,\ldots,r_n]$ &  \\
  &$\mid$& null $\;\;\;\mid\;\;\;$ $a$ & address & 
       & $\mid$& $\text{unpack}\ e_1\ \text{as}\ id : T[r_1,\ldots,r_n]\ \text{in}\ e_2$ &  \\
  &$\mid$& $\{ (a_i, iv), \ldots \}$ & coloring & 
    & & & \\
  &$\mid$& $\langle \langle \oton{\rho}, v\rangle \rangle$ & reg. relation instance & 
   & & & \\
%$bv$ & ::= & false $\;\;\;\mid\;\;\;$ true \\
%\\
%$iv$ & ::= & 0 $\;\;\;\mid\;\;\;$ 1 $\ldots$ \\
%\\
%$e$ & ::= & & expressions \\
%  &$\mid$& $bv$ $\;\;\;\mid\;\;\;$ $iv$ & constants \\
%  &$\mid$& $\langle e_1, e_2 \rangle$ $\;\;\;\mid\;\;\;$ $e$.1 $\;\;\;\mid\;\;\;$ $e$.2 & tuple \\
%  &$\mid$& $id$ &  \\
%  &$\mid$& $\text{new}\ T@r$ $\;\;\;\mid\;\;\;$ $\text{null }T@r$ $\;\;\;\mid\;\;\;$ $\text{isnull}(e)$ & \\
%  &$\mid$& $\text{upregion}(e, r_1,\ldots,r_n)$ $\;\;\;\mid\;\;\;$ $\text{downregion}(e, r_1,\ldots,r_n)$ & \\
%  &$\mid$& $\text{read}(e_1)$ $\;\;\;\mid\;\;\;$ $\text{write}(e_1, e_2)$ $\;\;\;\mid\;\;\;$ $\text{reduce}(id, e_1, e_2)$ & memory access \\
%  &$\mid$& $\text{newcolor}\ r$ $\;\;\;\mid\;\;\;$ $\text{color}(e_1, e_2, e_3)$ & coloring operations \\
%  &$\mid$& $e_1 + e_2$ & integer operations \\
%  &$\mid$& $e_1 < e_2$ & comparison operations \\
%  &$\mid$& $\text{let}\ id : T = e_1 \text{in}\ e_2$ &  \\
%  &$\mid$& $\text{if}\ e_1\ \text{then}\ e_2\ \text{else}\ e_3$ &  \\
%  &$\mid$& $id[r_1, \ldots, r_n](e_1,\ldots,e_n)$ & function calls \\
%  &$\mid$& $\text{partition}\ r_p\text{ using }e_1\text{ as }\oton{r}\text{ in }\ e_2$ &  \\
%  &$\mid$& $\text{pack}\ e_1\ \text{as}\ T[r_1,\ldots,r_n]$ &  \\
%  &$\mid$& $\text{unpack}\ e_1\ \text{as}\ id : T[r_1,\ldots,r_n]\ \text{in}\ e_2$ &  \\
\end{tabular}
}
\caption{Core Legion}
\label{fig:langdef}
\vspace{-5mm}
\end{figure*}

@@

\section{Legion}
\label{sec:legioncore}

@@

\subsection{Core Type System}
\label{subsec:coretypes}

Core Legion is explicitly typed using judgments of the form
$$\typeenv{e}{T}$$
where:
\begin{itemize}
\item $\Gamma$ is the type environment, mapping the names of local variables and global functions
to their types,
\item $\Phi$ is the region access privileges under which the expression is evaluated, and
\item $\Omega$ is the set of constraints that must hold between regions
\end{itemize}
As regions are hierarchical, privileges and/or constraints on a region often imply similar
privileges or constraints on subregions.  Figure~\ref{fig:closure} defines closure
operations for $\Phi^*$ and $\Omega^*$ that explicitly add these implied constraints.
The type checking rules are in Figure~\ref{fig:types}.  Many of the rules are 
straightforward, but several deserve more explanation.

@@

\label{fig:types}

@@

\begin{figure}
\centering{
$\begin{array}{l}
\Omega \subseteq \Omega^* \\
r_i \leq r_j \in \Omega^*  \Rightarrow r_i \leq r_i \in \Omega^*\wedge r_j \leq r_j \in \Omega^* \\
r_i \leq r_j \in \Omega^* \wedge r_j \leq r_k \in \Omega^* \Rightarrow r_i \leq r_k \in \Omega^* \\
r_i \leq r_j \in \Omega^* \wedge r_j * r_k \in \Omega^* \Rightarrow r_i * r_k \in \Omega^* \\
r_i * r_j \in \Omega^* \Rightarrow r_j * r_i \in \Omega^* \\
\\
\Phi \subseteq \Phi^* \\
r_i \leq r_j \in \Omega^* \wedge reads(r_j) \in \Phi^* \Rightarrow reads(r_i) \in \Phi^* \\
r_i \leq r_j \in \Omega^* \wedge writes(r_j) \in \Phi^* \Rightarrow writes(r_i) \in \Phi^* \\
r_i \leq r_j \in \Omega^* \wedge reduces_{id}(r_j) \in \Phi^* \Rightarrow reduces_{id}(r_i) \in \Phi^* \\
reads(r) \in \Phi^* \wedge writes(r) \in \Phi^* \Rightarrow reduces_{id}(r) \in \Phi^* \\
\ \ \ \mbox{for every function identifier $id$}
\end{array}$
}
\caption{Privilege and Constraint Closure}
\vspace{-5mm}
\label{fig:closure}
\end{figure}

\subsection{Operational Semantics}
\label{subsec:opsemantics}

@@

\item $valid\_interleave(S, C, E', \oton{E})$ is true if $E'$ is a valid interleaving
of the memory traces $\oton{E}$.
\end{itemize}
Figure~\ref{fig:opsemfns} defines {\em apply} and {\em mark\_coherence}.  The precise definition of
{\em valid\_interleave} is given in Section~\ref{sec:coherence}.  For now, we need only that a valid interleaving
is at least an interleaving, containing each constituent trace as a subsequence and no other operations.

@@

\begin{figure*}
\centering{
$
\begin{array}{@{}l@{}}
\begin{array}{@{}l@{\hspace{-0.25in}}l@{}} 
\begin{array}{@{}lll}
apply(S, []) & = & S \\
apply(S, E \dplus [ read(l, c, v) ]) & = & apply(S, E) \\
apply(S, E \dplus [ write(l, c, v) ]) & = & apply(S, E)[v/l] \\
apply(S, E \dplus [ reduce_{id}(l, c, v) ]) & = & S'[id(S'(l), v)/l], \\
& & \text{ where } S' = apply(S, E)
\end{array} &
\begin{array}{@{}ll}
mark\_coh([], \hat Q) & = [] \\
mark\_coh([ op(l, c, v) ] \dplus E, \hat Q) & = [ op(l, c', v) ] \dplus mark\_coh(E, \hat Q), \\
\multicolumn{2}{@{}l}{ \hspace{1.0in} \text{ where } c' =
\begin{cases}
simult, & \text{if }\exists \rho. l \in \rho \wedge simult(\rho) \in \hat Q \\
atomic, & \text{if }\exists \rho. l \in \rho \wedge atomic(\rho) \in \hat Q \\
excl, & \text{otherwise}
\end{cases}}
\end{array}
\end{array} \\
\begin{array}{@{}lll}
any\_interleave([], [], \ldots, []) & = & true \\
any\_interleave([ \epsilon ] \dplus E', E_1, \ldots, [ \epsilon ] \dplus E_i, \ldots, E_n) & = & any\_interleave(E', E_1, \ldots, E_i, \ldots, E_n) \\
valid\_interleave(S, C, E', \oton{E}) & \Rightarrow & any\_interleave(E', \oton{E}) \\
\end{array}
\end{array}
$}
\caption{Operational Semantics Helper Functions}
\label{fig:opsemfns}
\vspace{-5mm}
\end{figure*}

@@

\label{fig:semantics}

@@

\label{fig:semantics}

@@

\section{Soundness of Privileges}
\label{sec:soundness}

@@

\begin{figure*}
\centering{\footnotesize
$
\begin{array}{l}
\begin{array}{ll}
\begin{array}{lll}
\mapconsist{\Omega} & \Leftrightarrow & (\forall r_i, r_j. r_i \leq r_j \in \Omega \Rightarrow M(r_i) \subseteq M(r_j))\ \wedge \\
              &                 & (\forall r_i, r_j. r_i * r_j \in \Omega \Rightarrow M(r_i) \cap M(r_j) = \emptyset) \\
\localconsist{L}{\Gamma} & \Leftrightarrow & \forall (id,v) \in L. \valueconsist[]{v}{M \llbracket \Gamma \rrbracket (id)} \\
\storeconsist{S}{H} & \Leftrightarrow & \forall (l,v) \in S. \valueconsist[]{v}{H(l)} \\
\end{array} & 
\begin{array}{lll}
\traceconsist{E} & \Leftrightarrow & (\forall l, c, v. write(l,c, v) \in E \Rightarrow \valueconsist[]{v}{H(l)})\ \wedge \\
& & (\forall id, l, v. reduce_{id}(l, v) \in E \Rightarrow \\
& & (M \llbracket \Gamma \rrbracket (id) = (\hat T_1, \hat T_2), \emptyset, \emptyset \rightarrow \hat T_1) \wedge H(l) = \hat T_1 \wedge \valueconsist[]{v}{\hat T_2}) \\
\end{array}
\end{array} \vspace{3mm} \\
\begin{array}{l@{\hspace{1cm}}l}
\begin{array}{l}
\valueconsist[]{bv}{bool} \\
\valueconsist[]{iv}{int} \\
\valueconsist[]{null}{\hat T@\rho} \\
\end{array} &
\begin{array}{lll}
\valueconsist[]{l}{\hat T@\rho} & \Leftrightarrow & l \in \rho \wedge H(l) = \hat T \\
\valueconsist[]{\tuple{v_1}{v_2}}{\tuple{\hat T_1}{\hat T_2}} & \Leftrightarrow & (\valueconsist[]{v_1}{\hat T_1}) \wedge (\valueconsist[]{v_2}{\hat T_2}) \\
\valueconsist[]{\langle \langle \oton{\rho}, v \rangle \rangle}{\text{rr}[\oton{r}]\ \hat T\text{ where }\hat \Omega} & \Leftrightarrow & (\valueconsist[]{v}{\hat T[\rho_1/r_1, \ldots \rho_n/r_n]}) \wedge (\mapconsist[\{ (r_i, \rho_i) \}]{\hat \Omega}) \\
\valueconsist[]{K}{coloring(\rho)} & \Leftrightarrow & \forall l_1, v_1. (l_1, v_1) \in K \Rightarrow (l_1 \in \rho\ \wedge \\
& & \forall l_2, v_2. (l_2, v_2) \in K \Rightarrow (l_1 \not= l_2) \vee (v_1 = v_2)) \\
\end{array}
\end{array}
\end{array}
$
}
\caption{Consistency Properties}
\label{fig:constprop}
\vspace{-5mm}
\end{figure*}

\begin{thm}
\label{thm:effects}
\rm
If $\typeenv{e}{T}$ and $\opsenv{e}{v}{E}$  and $\mapconsist{\Omega}$, $\localconsist{L}{\Gamma}$ and $\storeconsist{S}{H}$,
then
$\valueconsist{v}{T}$, $\traceconsist{E}$ and $\privconsist{E}{\Phi}$.
\end{thm}

@@

\section{Coherence}
\label{sec:coherence}

Theorem~\ref{thm:effects} guarantees that a function call produces a value of the
correct type and does so within the regions of memory for which privileges are held.
It does not, however, address {\em coherence}, what concurrent operations on the regions are permitted.
An innovation in Legion is that programmers can control coherence on a per-region basis.  Intuitively,
the different coherence modes permit different interleavings of memory traces.  For example,
in addition to being an interleaving of the constituent traces, exclusive access requires interleaved memory traces to be:
\begin{itemize}
\item {\em coherent} - a read of location $l$ should see the effect of all writes and reductions to $l$
occurring earlier in the interleaved trace, and
\item {\em sequentially equivalent} - the value read by each read 
operation and the result of applying the interleaved trace to the original store should be the same as if each constituent trace
was applied to the store in a valid sequential execution of the program.
\end{itemize}
These requirements are relaxed for either a non-empty clobber set (representing
operations that may occur concurrently as part of evaluating other functions) or for coherence
modes other than {\em excl}.

>>>>

\setcounter{thm}{0}
\setcounter{lem}{0}
\texcomment{
\newtheorem{thm}{Theorem}
\newtheorem{lem}{Lemma}

@@

% fun latex tricks to make typeenv and opsenv more friendly
\makeatletter
\define@key{typeenv}{G}{\def\typeenv@G{#1}}
\define@key{typeenv}{P}{\def\typeenv@P{#1}}
\define@key{typeenv}{O}{\def\typeenv@O{#1}}
\newcommand{\typeenvx}[1][]{
{
% default values
\def\typeenv@G{\Gamma}
\def\typeenv@P{\Phi}
\def\typeenv@O{\Omega}
\setkeys{typeenv}{#1}
\typeenv@G, \typeenv@P, \typeenv@O \vdash \,
}}
\newcommand{\typeenv}[3][]{\typeenvx[#1] {#2} : {#3}}
\define@key{opsenv}{M}{\def\opsenv@M{#1}}
\define@key{opsenv}{L}{\def\opsenv@L{#1}}
\define@key{opsenv}{H}{\def\opsenv@H{#1}}
\define@key{opsenv}{S}{\def\opsenv@S{#1}}
\define@key{opsenv}{C}{\def\opsenv@C{#1}}
\newcommand{\opsenvx}[1][]{
{
% default values
\def\opsenv@M{M}
\def\opsenv@L{L}
\def\opsenv@H{H}
\def\opsenv@S{S}
\def\opsenv@C{C}
\setkeys{opsenv}{#1}
\opsenv@M, \opsenv@L, \opsenv@H, \opsenv@S, \opsenv@C \vdash \,
}}
\newcommand{\opsenv}[4][]{\opsenvx[#1] {#2} \mapsto {#3}, {#4}}
\makeatother
}

\texcomment{
\begin{figure*}
\centering
{\small
\begin{tabular}{cclr|cclr}

$T$ & ::= &  & types & $bv$ & ::= & false $\;\;\;\mid\;\;\;$ true & \\
  &$\mid$& bool $\;\;\;\mid\;\;\;$ int & base types & & & & \\
  &$\mid$& $\langle T_1, T_2 \rangle$ & tuple & $iv$ & ::= & 0 $\;\;\;\mid\;\;\;$ 1 $\ldots$ & \\
  &$\mid$& $T@(\oton{r})$ & pointer & & & & \\
  &$\mid$& $\text{coloring}(r)$ & region coloring & $e$ & ::= & & expressions \\
  &$\mid$& $\exists \oton{r}. T\text{ where }\Omega$ & region relationship &   &$\mid$& $bv$ $\;\;\;\mid\;\;\;$ $iv$ & constants \\
  &$\mid$& $\forall \oton{r}. (\oton{T}), \Phi, Q \rightarrow T_r$ & functions &   &$\mid$& $\langle e_1, e_2 \rangle$ $\;\;\;\mid\;\;\;$ $e$.1 $\;\;\;\mid\;\;\;$ $e$.2 & tuple \\
& & & &   &$\mid$& $id$ &  \\
$\Omega$ & ::= & $\{ \oton{\omega} \}$ & region constraints &   &$\mid$& $\text{new}\ T@r$ $\;\;\;\mid\;\;\;$ $\text{null }T@r$ $\;\;\;\mid\;\;\;$ $\text{isnull}(e)$ & \\
$\omega$ & ::= & $r_1 \leq r_2$ & subregion   &   
  &$\mid$& $\text{upregion}(e, r_1,\ldots,r_n)$ & \\
  &$\mid$& $r_1 * r_2$ & disjointness &   
  &$\mid$ & $\text{downregion}(e, r_1,\ldots,r_n)$ & \\  
& & &  & 
  &$\mid$& $\text{read}(e_1)$ $\;\mid\;$ $\text{write}(e_1, e_2)$ & memory access \\
$\Phi$ & ::= & $\{ \oton{\phi} \}$ & privileges & 
  & $\mid$ & $\text{reduce}(id, e_1, e_2)$ & \\
$\phi$ & ::= & reads$(r)$ $\;\mid\;$ writes$(r)$ $\;\mid\;$ $\text{reduces}_{id}(r)$ &  &  
  &$\mid$& $\text{newcolor}\ r$ $\;\mid\;$ $\text{color}(e_1, e_2, e_3)$ & coloring \\
& & & &  
  &   $\mid$& $e_1 + e_2$ & integer ops \\
$Q$ & ::= & $\{ \oton{q} \}$ & coherence modes  &   
       & $\mid$& $e_1 < e_2$ & comparisons \\
$q$ & ::= & atomic$(r)$ $\;\;\;\mid\;\;\;$ simult$(r)$ & &   
  &$\mid$& $\text{let}\ id : T = e_1 \text{in}\ e_2$ &  \\
& & & &   
       & $\mid$& $\text{if}\ e_1\ \text{then}\ e_2\ \text{else}\ e_3$ &  \\
$v$ & ::= & & values &   
       & $\mid$& $id[r_1, \ldots, r_n](e_1,\ldots,e_n)$ & function calls \\
  &$\mid$& $bv$ $\;\;\;\mid\;\;\;$ $iv$ & base values &   
       & $\mid$ & $\text{partition}\ r_p\text{ using }e_1\text{ as }\oton{r}\text{ in }\ e_2$ &  \\
  &$\mid$& $\langle v_1, v_2 \rangle$ & tuple & 
       & $\mid$& $\text{pack}\ e_1\ \text{as}\ T[r_1,\ldots,r_n]$ &  \\
  &$\mid$& null $\;\;\;\mid\;\;\;$ $a$ & address & 
       & $\mid$& $\text{unpack}\ e_1\ \text{as}\ id : T[r_1,\ldots,r_n]\ \text{in}\ e_2$ &  \\
  &$\mid$& $\{ (a_i, iv), \ldots \}$ & coloring & 
    & & & \\
  &$\mid$& $\langle \langle \oton{\rho}, v\rangle \rangle$ & reg. relation instance & 
   & & & \\
%$bv$ & ::= & false $\;\;\;\mid\;\;\;$ true \\
%\\
%$iv$ & ::= & 0 $\;\;\;\mid\;\;\;$ 1 $\ldots$ \\
%\\
%$e$ & ::= & & expressions \\
%  &$\mid$& $bv$ $\;\;\;\mid\;\;\;$ $iv$ & constants \\
%  &$\mid$& $\langle e_1, e_2 \rangle$ $\;\;\;\mid\;\;\;$ $e$.1 $\;\;\;\mid\;\;\;$ $e$.2 & tuple \\
%  &$\mid$& $id$ &  \\
%  &$\mid$& $\text{new}\ T@r$ $\;\;\;\mid\;\;\;$ $\text{null }T@r$ $\;\;\;\mid\;\;\;$ $\text{isnull}(e)$ & \\
%  &$\mid$& $\text{upregion}(e, r_1,\ldots,r_n)$ $\;\;\;\mid\;\;\;$ $\text{downregion}(e, r_1,\ldots,r_n)$ & \\
%  &$\mid$& $\text{read}(e_1)$ $\;\;\;\mid\;\;\;$ $\text{write}(e_1, e_2)$ $\;\;\;\mid\;\;\;$ $\text{reduce}(id, e_1, e_2)$ & memory access \\
%  &$\mid$& $\text{newcolor}\ r$ $\;\;\;\mid\;\;\;$ $\text{color}(e_1, e_2, e_3)$ & coloring operations \\
%  &$\mid$& $e_1 + e_2$ & integer operations \\
%  &$\mid$& $e_1 < e_2$ & comparison operations \\
%  &$\mid$& $\text{let}\ id : T = e_1 \text{in}\ e_2$ &  \\
%  &$\mid$& $\text{if}\ e_1\ \text{then}\ e_2\ \text{else}\ e_3$ &  \\
%  &$\mid$& $id[r_1, \ldots, r_n](e_1,\ldots,e_n)$ & function calls \\
%  &$\mid$& $\text{partition}\ r_p\text{ using }e_1\text{ as }\oton{r}\text{ in }\ e_2$ &  \\
%  &$\mid$& $\text{pack}\ e_1\ \text{as}\ T[r_1,\ldots,r_n]$ &  \\
%  &$\mid$& $\text{unpack}\ e_1\ \text{as}\ id : T[r_1,\ldots,r_n]\ \text{in}\ e_2$ &  \\
\end{tabular}
}
\caption{Core Legion}
\label{fig:langdeffull}
\vspace{-5mm}
\end{figure*}
}

@@

\section{Core Legion}
\label{sec:legioncorefull}

@@

\subsection{Core Type System}
\label{subsec:coretypesfull}

Core Legion is explicitly typed using judgments of the form
$$\typeenv{e}{T}$$
where:
\begin{itemize}
\item $\Gamma$ is the type environment, mapping the names of local variables and global functions
to their types,
\item $\Phi$ is the region access privileges under which the expression is evaluated, and
\item $\Omega$ is the set of constraints that must hold between regions
\end{itemize}
As regions are hierarchical, privileges and/or constraints on a region often imply similar
privileges or constraints on subregions.  Figure~\ref{fig:closure} defines closure
operations for $\Phi^*$ and $\Omega^*$ that explicitly add these implied constraints.
The type checking rules are in Figure~\ref{fig:typesfull}.  Many of the rules are 
straightforward, but several deserve more explanation.

@@

\label{fig:typesfull}

@@

\texcomment{
\begin{figure}
\centering{
$\begin{array}{l}
\Omega \subseteq \Omega^* \\
r_i \leq r_j \in \Omega^*  \Rightarrow r_i \leq r_i \in \Omega^*\wedge r_j \leq r_j \in \Omega^* \\
r_i \leq r_j \in \Omega^* \wedge r_j \leq r_k \in \Omega^* \Rightarrow r_i \leq r_k \in \Omega^* \\
r_i \leq r_j \in \Omega^* \wedge r_j * r_k \in \Omega^* \Rightarrow r_i * r_k \in \Omega^* \\
r_i * r_j \in \Omega^* \Rightarrow r_j * r_i \in \Omega^* \\
\\
\Phi \subseteq \Phi^* \\
r_i \leq r_j \in \Omega^* \wedge reads(r_j) \in \Phi^* \Rightarrow reads(r_i) \in \Phi^* \\
r_i \leq r_j \in \Omega^* \wedge writes(r_j) \in \Phi^* \Rightarrow writes(r_i) \in \Phi^* \\
r_i \leq r_j \in \Omega^* \wedge reduces_{id}(r_j) \in \Phi^* \Rightarrow reduces_{id}(r_i) \in \Phi^* \\
reads(r) \in \Phi^* \wedge writes(r) \in \Phi^* \Rightarrow reduces_{id}(r) \in \Phi^* \\
\ \ \ \mbox{for every function identifier $id$}
\end{array}$
}
\caption{Privilege and Constraint Closure}
\vspace{-5mm}
\label{fig:closure}
\end{figure}
}

\subsection{Operational Semantics}
\label{subsec:opsemanticsfull}

@@

\item $valid\_interleave(S, C, E', \oton{E})$ is true if $E'$ is a valid interleaving
of the memory traces $\oton{E}$.
\end{itemize}
Figure~\ref{fig:langdef} defines {\em apply} and {\em mark\_coherence}.  The precise definition of
{\em valid\_interleave} is given in Section~\ref{sec:coherencefull}.  For now, we need only that a valid interleaving
is at least an interleaving, containing each constituent trace as a subsequence and no other operations.

@@

\texcomment{
\begin{figure*}
\centering{
$
\begin{array}{@{}l@{}}
\begin{array}{@{}l@{\hspace{-0.25in}}l@{}} 
\begin{array}{@{}lll}
apply(S, []) & = & S \\
apply(S, E \dplus [ read(l, c, v) ]) & = & apply(S, E) \\
apply(S, E \dplus [ write(l, c, v) ]) & = & apply(S, E)[v/l] \\
apply(S, E \dplus [ reduce_{id}(l, c, v) ]) & = & S'[id(S'(l), v)/l], \\
& & \text{ where } S' = apply(S, E)
\end{array} &
\begin{array}{@{}ll}
mark\_coh([], \hat Q) & = [] \\
mark\_coh([ op(l, c, v) ] \dplus E, \hat Q) & = [ op(l, c', v) ] \dplus mark\_coh(E, \hat Q), \\
\multicolumn{2}{@{}l}{ \hspace{1.0in} \text{ where } c' =
\begin{cases}
simult, & \text{if }\exists \rho. l \in \rho \wedge simult(\rho) \in \hat Q \\
atomic, & \text{if }\exists \rho. l \in \rho \wedge atomic(\rho) \in \hat Q \\
excl, & \text{otherwise}
\end{cases}}
\end{array}
\end{array} \\
\begin{array}{@{}lll}
any\_interleave([], [], \ldots, []) & = & true \\
any\_interleave([ \epsilon ] \dplus E', E_1, \ldots, [ \epsilon ] \dplus E_i, \ldots, E_n) & = & any\_interleave(E', E_1, \ldots, E_i, \ldots, E_n) \\
valid\_interleave(S, C, E', \oton{E}) & \Rightarrow & any\_interleave(E', \oton{E}) \\
\end{array}
\end{array}
$}
\caption{Operational Semantics Helper Functions}
\label{fig:opsemfnsfull}
\vspace{-5mm}
\end{figure*}
}

@@

\label{fig:semanticsfull}

@@

\label{fig:semanticfull}

@@

\section{Soundness of Privileges}
\label{sec:soundnessfull}

@@

\texcomment{
\begin{figure*}
\centering{\footnotesize
$
\begin{array}{l}
\begin{array}{ll}
\begin{array}{lll}
\mapconsist{\Omega} & \Leftrightarrow & (\forall r_i, r_j. r_i \leq r_j \in \Omega \Rightarrow M(r_i) \subseteq M(r_j))\ \wedge \\
              &                 & (\forall r_i, r_j. r_i * r_j \in \Omega \Rightarrow M(r_i) \cap M(r_j) = \emptyset) \\
\localconsist{L}{\Gamma} & \Leftrightarrow & \forall (id,v) \in L. \valueconsist[]{v}{M \llbracket \Gamma \rrbracket (id)} \\
\storeconsist{S}{H} & \Leftrightarrow & \forall (l,v) \in S. \valueconsist[]{v}{H(l)} \\
\end{array} & 
\begin{array}{lll}
\traceconsist{E} & \Leftrightarrow & (\forall l, c, v. write(l,c, v) \in E \Rightarrow \valueconsist[]{v}{H(l)})\ \wedge \\
& & (\forall id, l, v. reduce_{id}(l, v) \in E \Rightarrow \\
& & (M \llbracket \Gamma \rrbracket (id) = (\hat T_1, \hat T_2), \emptyset, \emptyset \rightarrow \hat T_1) \wedge H(l) = \hat T_1 \wedge \valueconsist[]{v}{\hat T_2}) \\
\end{array}
\end{array} \vspace{3mm} \\
\begin{array}{l@{\hspace{1cm}}l}
\begin{array}{l}
\valueconsist[]{bv}{bool} \\
\valueconsist[]{iv}{int} \\
\valueconsist[]{null}{\hat T@\rho} \\
\end{array} &
\begin{array}{lll}
\valueconsist[]{l}{\hat T@\rho} & \Leftrightarrow & l \in \rho \wedge H(l) = \hat T \\
\valueconsist[]{\tuple{v_1}{v_2}}{\tuple{\hat T_1}{\hat T_2}} & \Leftrightarrow & (\valueconsist[]{v_1}{\hat T_1}) \wedge (\valueconsist[]{v_2}{\hat T_2}) \\
\valueconsist[]{\langle \langle \oton{\rho}, v \rangle \rangle}{\text{rr}[\oton{r}]\ \hat T\text{ where }\hat \Omega} & \Leftrightarrow & (\valueconsist[]{v}{\hat T[\rho_1/r_1, \ldots \rho_n/r_n]}) \wedge (\mapconsist[\{ (r_i, \rho_i) \}]{\hat \Omega}) \\
\valueconsist[]{K}{coloring(\rho)} & \Leftrightarrow & \forall l_1, v_1. (l_1, v_1) \in K \Rightarrow (l_1 \in \rho\ \wedge \\
& & \forall l_2, v_2. (l_2, v_2) \in K \Rightarrow (l_1 \not= l_2) \vee (v_1 = v_2)) \\
\end{array}
\end{array}
\end{array}
$
}
\caption{Consistency Properties}
\label{fig:constpropfull}
\vspace{-5mm}
\end{figure*}
}

\begin{thm}
\label{thm:effectsfull}
\rm
If $\typeenv{e}{T}$ and $\opsenv{e}{v}{E}$  and $\mapconsist{\Omega}$, $\localconsist{L}{\Gamma}$ and $\storeconsist{S}{H}$,
then
$\valueconsist{v}{T}$, $\traceconsist{E}$ and $\privconsist{E}{\Phi}$.
\end{thm}

@@

\section{Coherence}
\label{sec:coherencefull}

Theorem~\ref{thm:effectsfull} guarantees that a function call produces a value of the
correct type and does so within the regions of memory for which privileges are held.
It does not, however, address {\em coherence}, what concurrent operations on the regions are permitted.
An innovation in Legion is that programmers can control coherence on a per-region basis.  Intuitively,
the different coherence modes permit different interleavings of memory traces.  For example,
in addition to being an interleaving of the constituent traces, exclusive access requires interleaved memory traces to be:
\begin{itemize}
\item {\em coherent} - a read of location $l$ should see the effect of all writes and reductions to $l$
occurring earlier in the interleaved trace, and
\item {\em sequentially equivalent} - the value read by each read 
operation and the result of applying the interleaved trace to the original store should be the same as if each constituent trace
was applied to the store in a valid sequential execution of the program.
\end{itemize}
These requirements are relaxed for either a non-empty clobber set (representing
operations that may occur concurrently as part of evaluating other functions) or for coherence
modes other than {\em excl}.
